\chapter{Introduction}

Game playing applications are not a recent phenomenon and can be traced back to Arthur Samuel's Checkers Player \citep{samuelStudiesMachineLearning1959} that used heuristic search methodology and \acrfull{td} learning \citep[p.~121]{suttonReinforcementLearningIntroduction2020} together with linear function approximation. 

The game of backgammon, with its vastly larger state space and an effective branching factor of approximately 400 \citep[p.~422]{suttonReinforcementLearningIntroduction2020}, demanded a far higher representational capacity than linear function approximation could provide. \citet{tesauroTDGammonSelfTeachingBackgammon1994} addressed this challenge with TD-Gammon, a \acrfull{rl} program that required little backgammon knowledge and yet was able to learn to play at the level of the world's strongest players using TD($\lambda$) algorithm and non-linear function approximation using \acrfull{ann} \citep[p.~421]{suttonReinforcementLearningIntroduction2020}. The first version of TD-Gammon, using only raw board information as input, already outperformed programs trained with extensive backgammon knowledge \citep[p.~266]{tesauroPracticalIssuesTemporal1992}. This early success demonstrated the power of self-play and highlighted the merits of learning directly from experience. In TD-Gammon 1.0 \citet[p.~20]{tesauroTDGammonSelfTeachingBackgammon1994} improved the level of play to match that of human experts by using hand-crafted backgammon features.

More than two decades would pass before \citet[p.~529]{mnihHumanlevelControlDeep2015} introduced \acrfull{dqn}, an artificial agent capable of learning successful control policies directly from high-dimensional sensory inputs using \acrshort{rl} and \acrfull{dnn}. The agent required no hand-crafted features whatsoever. Using only raw pixels and game scores as input, \acrshort{dqn} exceeded all prior algorithms and achieved human-level performance comparable to a professional game tester across 49 Atari 2600 games \citep[Figure~3]{mnihHumanlevelControlDeep2015}. Remarkably, a single algorithm, network architecture, and set of hyperparameters sufficed for learning problem-specific features automatically for all games.

Pong was among the 49 Atari 2600 games that \acrshort{dqn} was tested on, leveraging the \acrfull{ale} \citep{bellemareArcadeLearningEnvironment2013} as a testbed for studying learning and planning algorithms. All games shared a common observation space of raw pixel inputs, preprocessed identically without any game-specific knowledge, yet they differed in their action spaces, state-transition dynamics, and the optimal policies required to maximize score \citep[p.~437]{suttonReinforcementLearningIntroduction2020}. This diverse yet standardized benchmark enabled meaningful comparison of different \acrshort{rl} algorithms. Among these, Pong - a simple game in which two paddles move up or down to hit a ball back and forth, with a reward of +1 for each point scored and -1 for each point conceded - became a recurring test environment across the major deep \acrshort{rl} architectures \citep{mnihHumanlevelControlDeep2015,mnihAsynchronousMethodsDeep2016,schulmanProximalPolicyOptimization2017}. Its simple rules, discrete action space, and sparse reward signal make it an effective testbed for comparing algorithms that take fundamentally different approaches to learning from high-dimensional visual input.

Begin Paragraph3 with the central question - pong was used in DQN as well as A3C - two different paradigms with very different tradeoffs. Introduce the fact that this paper deals with how these 2 paradigms compare using the game of Pong.

\textcolor{red}{Central question: DQN vs. A2C on Pong}

$\;$ \textcolor{red}{sample efficiency}

$\;$ \textcolor{red}{stability}

$\;$ \textcolor{red}{performance}

\textcolor{red}{Scope boundaries}

\textcolor{red}{Definitions}

\textcolor{red}{Structural outline}