\chapter {Policy-Based Deep Reinforcement Learning - From REINFORCE to A2C}
\label{chap:a2c}

This chapter develops the second path identified in Section~\ref{sec:policyoptimization}. Instead of approximating $q_*$ and deriving the policy via $\argmax$, the policy is directly parameterized as $\pi(a|s, \theta)$ and $\theta$ is optimized by gradient ascent on a scalar performance measure. A value function may still be used to evaluate and improve the policy, but it is not used for action selection.

\section{The Policy Gradient Theorem}
Besides optimizing the policy directly, parameterized policy methods offer two further advantages over action-value methods. First, a soft-max policy $\pi(a|s, \theta)$ \citep[p.~322]{suttonReinforcementLearningIntroduction2020} assigns a learned probability to every action, allowing the agent to discover that a mixture of actions outperforms any single greedy choice. An $\epsilon$-greedy policy cannot do this because it commits to one action with probability 1 - $\epsilon$ and distributes the remaining $\epsilon$ uniformly, regardless of what the agent has learned about non-greedy actions. Second, with a continuous policy parameterization, small changes in $\theta$ produce small smooth changes in the action probabilities. With $\epsilon$-greedy action selection over action values, an arbitrarily small change in the estimated values can cause the $\argmax$ to switch from one action to another, producing a discontinuous change in the policy \citep[p.~324]{suttonReinforcementLearningIntroduction2020}.

For the episodic case, such as Pong, the policy parameter $\theta$ is optimized by a gradient ascent on a scalar performance measure defined as the value of the start state:
\begin{equation}
	J(\theta) \doteq v_{\pi_{\theta}}(s_0),
\end{equation}
where $v_{\pi_{\theta}}$ is the true value function for $\pi_{\theta}$ determined by $\theta$, and $s_0$ is the start state. The gradient $\nabla J(\theta)$, however, is not straightforward to compute since changing $\theta$ changes both the action selection and the state distribution that the policy induces \citep[p.~324]{suttonReinforcementLearningIntroduction2020}. Computing the effect of the policy on the state distribution requires the environment dynamics, which are typically unknown \citep[p.~325]{suttonReinforcementLearningIntroduction2020}. A solution to this is provided by the policy gradient theorem:
\begin{equation}\label{eq:policy_gradient_theorem}
	\nabla J(\theta) \propto \sum_s \mu(s) \sum_a q_\pi(s, a)\,\nabla \pi(a\,|\,s, \theta),
\end{equation}
where $\mu(s)$ is the on-policy state distribution that appears only as a weighting and does not need to be differentiated. By sampling transitions through interaction with the environment, the agent visits states according to $\mu(s)$ naturally, which allows the gradient to be estimated without knowledge of the environment dynamics.

\section{REINFORCE and the Variance Problem}
The policy gradient theorem contains $ q_\pi(s, a)$ which requires the expectation over all future rewards under $\pi$ and is therefore unavailable in closed form. A gradient estimator can be obtained in two steps. First, the sum over actions is rewritten using the log-likelihood trick. Second, the unknown $q_\pi$ is replaced by a sample return $G_t$ from \eqref{eq:action_value_function}. Applying both steps to \eqref{eq:policy_gradient_theorem} gives
\begin{align}
	\nabla J(\theta) &= \mathbb{E}_\pi \bigg[\sum_a q_\pi(S_t, a)\,\nabla \pi(a\,|\,S_t, \theta)\bigg] \nonumber \\
	&= \mathbb{E}_\pi \bigg[q_\pi(S_t, A_t) \frac{\nabla \pi(A_t\,|\,S_t, \theta)}{\pi(A_t\,|\,S_t, \theta)} \bigg] \nonumber \\
	&=  \mathbb{E}_\pi \bigg[G_t \nabla \ln \pi(A_t\,|S_t, \theta) \bigg].
\end{align}
This final expression is the REINFORCE gradient estimator \citep[p.~327]{williamsSimpleStatisticalGradientfollowing1992, suttonReinforcementLearningIntroduction2020}. It requires only quantities available from a single trajectory, namely the return $G_t$, the state $S_t$, the action taken $A_t$, and the gradient of the log-policy.

Since $G_t$ is a full Monte Carlo return, it accumulates the randomness of every subsequent action and state transition from timestep $t$ to the end of the episode. In Pong, where episodes run for hundreds of frames and rewards are sparse, two trajectories starting from the same state can produce vastly different returns simply because the stochastic policy may select different actions at each subsequent timestep. The resulting gradient estimate $\nabla J(\theta)$ is correct in expectation but extremely noisy, requiring many episodes to converge.

\citet[p.~329]{suttonReinforcementLearningIntroduction2020} shows that subtracting a state dependent baseline $b(S_t)$ \citep{williamsSimpleStatisticalGradientfollowing1992} from $G_t$ does not change the expected gradient but reduces its variance. The modified gradient estimator becomes
\begin{equation}\label{eq:reinforce_baseline}
	\nabla J(\theta) = \mathbb{E}_\pi \bigg[(G_t - b(S_t))\, \nabla \ln \pi(A_t\,|\,S_t, \theta) \bigg].
\end{equation}
When $b(S_t) \approx v_\pi(S_t)$, the quantity $(G_t - b(S_t))$ measures how much better a specific trajectory was compared to the average from state $S_t$. This centers the gradient signal around zero rather than around the potentially large absolute return, reducing the magnitude of fluctuations across episodes. However, the baseline does not address the root cause of the variance. The return $G_t$ remains a full Monte Carlo estimate that depends on all future rewards from timestep $t$ onward. Section~\ref{sec:actorcriticmethods} resolves this by replacing $G_t$ with a bootstrapped target.

\section{Actor-Critic Methods}
\label{sec:actorcriticmethods}
The same bootstrapping principle introduced for the action-value function in Section~\ref{sec:fromqlearningtodqn} can be applied here. Rather than waiting until the end of the episode to compute $G_t$, the return can be approximated by the immediate reward and an estimate of the next state's value, giving the one-step \acrshort{td} target $R_{t+1} + \gamma \hat{V}(S_{t+1}, w)$. \citet[p.~331]{suttonReinforcementLearningIntroduction2020} formulates the one-step update in the actor-critic methods as follows:
\begin{equation}\label{eq:one_step_update}
	\theta_{t+1} \doteq \theta_t + \alpha_{\theta}\,\delta_t \nabla \ln \pi(A_t\,|\,S_t, \theta),
\end{equation}
where $\delta_t = R_{t+1} + \gamma \hat{V}(S_{t+1}, w) - \hat{V}(S_t, w)$ is the \acrshort{td} error. Since the \acrshort{td} error depends only on the immediate reward and the value estimates at two consecutive states, it no longer accumulates the variance of all future rewards. The resulting gradient estimate has vastly lower variance but introduces a bias because $\hat{V}$ is a learned approximation that may differ from the true value function $v_\pi$.

The actor-critic method implied by equation~\eqref{eq:one_step_update} has two learnable components. The actor is the policy $\pi(a\,|\,s, \theta)$, parameterized by $\theta$, which decides which action to take. The critic is the value function $\hat{V}(s, w)$, parameterized by w, which estimates the expected return from each state under the current policy. The critic's update is a semi-gradient \acrshort{td} \citep[p.~203]{suttonReinforcementLearningIntroduction2020}:
\begin{equation}
	w_{t+1} \doteq w_t + \alpha_{w}\,\delta_t \nabla \hat{V}(S_t, w),
\end{equation}
where the gradient is taken only with respect to $\hat{V}(S_t, w)$ and the bootstrapped target is treated as a constant.

The actor-critic architecture differs structurally from \acrshort{dqn}. In \acrshort{dqn} the neural network is the entire decision-maker. It outputs $Q$-values, and the policy falls out via $\argmax$. In actor-critic architecture the network splits into an actor and a critic component each with a distinct role. The critic supports the actor but does not directly determine actions. Section~\ref{sec:algandstructdifferences} compares these structural differences in detail.


\section{The Advantage Function and A2C}
\label{sec:advantagefunctionanda2c}

The \acrshort{td} error $\delta_t$ used in section~\ref{sec:actorcriticmethods} is actually an estimate of a specific quantity known as the advantage function \citep{mnihAsynchronousMethodsDeep2016},
\begin{equation}
	A_\pi(s, a) = q_\pi(s, a) - v_\pi(s).
\end{equation}
 
 It measures how much better action $a$ is compared to the average over all actions available in state $s$ under $\pi$. If $A > 0$, the action is above average and should be reinforced. If  $A < 0$, it is below average and should be suppressed. This is more informative than using $q_\pi$ alone because it adjusts for the state's baseline value. 
 
The REINFORCE update with baseline from \eqref{eq:reinforce_baseline} was already estimating this quantity implicitly. The sample return $G_t$ is an estimate of $q_\pi(S_t, A_t)$ and the baseline $b(S_t)$ is an estimate of $v_\pi(S_t)$, making the difference $(G_t - b(S_t))$ a Monte Carlo estimate of the advantage \citep{mnihAsynchronousMethodsDeep2016}. The actor-critic framework of Section~\ref{sec:actorcriticmethods} replaces this Monte Carlo estimate with the TD error $\delta_t = R_{t+1} + \gamma \hat{V}(S_{t+1}, w) - \hat{V}(S_t, w)$ which estimates the same quantity. In expectation:
\begin{align}
	\mathbb{E}[\delta_t\,|\, S_t = s, A_t = a] &= \mathbb{E}[R_{t+1} + \gamma\,v_\pi(S_{t+1})\,|\,s, a] - v_\pi(s) \nonumber \\
	&= q_\pi(s,a) - v_\pi(s) \nonumber \\
	&= A_\pi(s,a)
\end{align}

REINFORCE with baseline and actor-critic are therefore not two unrelated algorithms but two estimators of the same underlying quantity. They differ only in how they estimate the advantage where the former uses a full Monte Carlo return with high variance and no bias, while the latter uses a bootstrapped TD target with lower variance at the cost of bias from the learned value function.

The advantage actor-critic framework formalizes this and adds two practical enhancements that make the method viable on visually complex environment like Pong. The first is synchronous parallel data collection. Instead of training on transitions from a single environment, \acrshort{a2c} runs $N$ copies of the environment simultaneously, each executing the current policy $\pi(\cdot\,|\,\cdot, \theta)$ independently. At each update step, the agent collects short rollouts of length $n$ from all $N$ environments, computes the advantage estimate $\delta_t$ for each transition, and averages the resulting gradients. Averaging over $N$ independent gradient estimates reduces the variance of the update by a factor of approximately $\sqrt{N}$, partially compensating for the on-policy requirement that each transition can only be used once. The synchronous design, in contrast to the asynchronous variant \acrshort{a3c} \citep{mnihAsynchronousMethodsDeep2016}, avoids stale gradients by ensuring all environments use the same policy parameters at each update.

The second enhancement is entropy regularization. Without explicit encouragement to explore, the policy can collapse to a near-deterministic distribution early in training, before the critic has learned accurate value estimates. To prevent this, \acrshort{a2c} adds the entropy of the policy $H(\pi(\cdot\,|\,S_t, \theta))$ to the objective, a technique originally proposed by \citet{williamsFunctionOptimizationUsing1991} and adopted in the \acrshort{a3c} framework by \citet{mnihAsynchronousMethodsDeep2016}. The combined loss for the actor becomes
\begin{equation}\label{eq:a2c_actor_loss}
	L_\text{actor}(\theta) = -\mathbb{E}_\pi\Big[\delta_t \ln \pi(A_t\,|\,S_t, \theta) + \beta\, H\big(\pi(\cdot\,|\,S_t, \theta)\big)\Big],
\end{equation}
where $\beta$ controls the strength of the regularization. A high entropy means the policy spreads probability across multiple actions, encouraging exploration. As training progresses and the critic's estimates improve, the policy naturally sharpens toward higher-advantage actions while the entropy term prevents premature commitment. This serves a similar purpose to $\epsilon$-greedy exploration in DQN but arises from the policy parameterization itself rather than being imposed externally.
In practice, the actor and critic share a convolutional feature extraction backbone and split into two output heads, a softmax layer producing the policy $\pi(a\,|\,s, \theta)$ and a single linear output estimating $\hat{V}(s, w)$. Sharing the lower layers is efficient because both components need to extract the same visual features from the raw pixel input. The architectural details for Pong are presented in Section~\ref{sec:a2c_results}.

Chapters \ref{chap:dqn} and \ref{chap:a2c} have developed two distinct approaches to the same problem. \acrshort{dqn} learns an action-value function and extracts the policy implicitly through an argmax operation, producing a deterministic mapping from states to actions. \acrshort{a2c} learns the policy explicitly as a probability distribution over actions, guided by a critic that evaluates each decision through the advantage function. Both methods bootstrap from learned value estimates, but they differ in what they bootstrap (Q-values versus state values), how they collect experience (off-policy replay versus on-policy parallel rollouts), and how they explore (externally imposed $\epsilon$-greedy decay versus entropy regularization intrinsic to the policy). Chapter \ref{chap:pong} applies both methods to the same Pong environment, providing the empirical basis for a direct comparison along these axes.
